---
title: "AunerLab_CCLE"
author: "Kevin Blighe"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
fig_width: 7
vignette: >
    %\VignetteIndexEntry{AunerLab_CCLE}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---


# Set-up

```{r, echo=FALSE}

  library(knitr)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

```

## install and load necessary packages

```{r, echo=TRUE}

  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')

  if (!requireNamespace('ggplot2', quietly = TRUE))
    BiocManager::install('ggplot2')

  if (!requireNamespace('reshape2', quietly = TRUE))
    BiocManager::install('reshape2')

  if (!requireNamespace('data.table', quietly = TRUE))
    BiocManager::install('data.table')

  if (!requireNamespace('gridExtra', quietly = TRUE))
    BiocManager::install('gridExtra')

  if (!requireNamespace('grid', quietly = TRUE))
    BiocManager::install('grid')

  library(ggplot2)
  library(grid)
  library(gridExtra)

```

## download scripts

Download the 2 functions in the '*R*' directory on this Github repo, either directly or via `git clone`, and ensure that these are placed in a directory called '*R*' in your current working directory, i.e.:

 - R/ccle_barplot.R

 - R/ccle_correlation.R


*HINT: to check what is your current working directory, type `getwd()`. It can also be set to something different via `setwd()`*

## download data

Download the files *CCLE_RNAseq_genes_rpkm_20180929.gct.gz* and *Cell_lines_annotations_20181226.txt* from https://portals.broadinstitute.org/ccle/data (sign in required), and place them into a directory called '*data*' in your current working directory, i.e.:

 - data/CCLE_RNAseq_genes_rpkm_20180929.gct.gz
 
 - data/Cell_lines_annotations_20181226.txt



# Prepare data

## read in and tidy expression data

```{r read_expression_data}

  ccle <- data.table::fread(
    "gunzip -c data/CCLE_RNAseq_genes_rpkm_20180929.gct.gz",
    header = TRUE,
    sep = '\t',
    stringsAsFactors = FALSE,
    skip = 2,
    data.table = FALSE)

  # set rownames
  rownames(ccle) <- make.unique(ccle$Description)

  # remove first 2 columns to have only numerical data
  ccle <- ccle[,3:ncol(ccle)]

  # tidy up colnames
  colnames(ccle) <- gsub("^X", "", gsub("\\.\\.ACH\\.[0-9]*\\.$", "", colnames(ccle)))

```


## read in metadata for cell-lines

```{r read_metadata}

  sampleinfo <- read.csv(
    "data/Cell_lines_annotations_20181226.txt",
    header = TRUE,
    sep = "\t",
    stringsAsFactors = FALSE,
    skip = 0)

```

## view the different histological types

```{r view_cell_line_types}

  sort(unique(sampleinfo$Hist_Subtype1))

```

## check numbers of cell-lines in each category

```{r check_cell_line_numbers}

  sort(table(sampleinfo$Hist_Subtype1))

```

### extract names of plasma cell myeloma / multiple myeloma lines

```{r extract_MM_cell_lines}

  MMlines <- sampleinfo[grep("plasma_cell_myeloma", sampleinfo$Hist_Subtype1),1]

```

### which MM lines are not in the expression data?

```{r MM_cell_lines_not_in_expression_data}

  MMlines[-which(MMlines %in% colnames(ccle))]

```


### filter expression data to only include multiple myeloma lines

```{r filter_expression_data}

  MMccle <- ccle[,which(colnames(ccle) %in% MMlines)]

  colnames(MMccle)

  MMccle[1:5,1:5]

```




# Visualise data

## barplot

Read the barplot function code:

```{r read_barplot_code}

  source('R/ccle_barplot.R')

```


Now, generate 3 barplots:

```{r generate_barplot}

  p1 <- ccle_barplot(
    ccledata = ccle,
    clinicaldata = sampleinfo,
    keyword = 'Hodgkin_lymphoma',
    gene = 'TP53',
    title = 'Hodgkin\'s lymphoma',
    xlab = 'Cell-line',
    greyscale = TRUE,
    colour = NULL,
    titlesize = 18,
    axissize = 14)
  
  p2 <- ccle_barplot(
    ccledata = ccle,
    clinicaldata = sampleinfo,
    keyword = 'chronic_myeloid_leukaemia',
    gene = 'TP53',
    title = 'Chronic myeloid leukaemia',
    xlab = 'Cell-line',
    greyscale = FALSE,
    titlesize = 18,
    axissize = 14)

  p3 <- ccle_barplot(
    ccledata = ccle,
    clinicaldata = sampleinfo,
    keyword = 'immortalized_epithelial',
    gene = 'TP53',
    title = 'Immortalized epithelial',
    xlab = 'Cell-line',
    greyscale = FALSE,
    colour = 'royalblue',
    titlesize = 18,
    axissize = 14)

```

Now, plot these together:

```{r plot_barplot, fig.width = 15, fig.height = 6}

  grid.arrange(p1, p2, p3, ncol = 3, nrow = 1,
    top = textGrob('Title', gp = gpar(fontsize = 24, fontface="bold")))

  grid.rect(gp = gpar(fill = NA))

```

## correlation plot and analysis

Read the correlation function code:

```{r read_correlation_code}

  source('R/ccle_correlation.R')

```


Now, generate 2 correlation plots:

```{r generate_correlation}

  p1 <- ccle_correlation(
    ccledata = ccle,
    clinicaldata = sampleinfo,
    keyword = 'Hodgkin_lymphoma',
    gene1 = 'TP53',
    gene2 = 'ATM',
    title = 'Hodgkin\'s lymphoma',
    titlesize = 24,
    axissize = 16)

  p2 <- ccle_correlation(
    ccledata = ccle,
    clinicaldata = sampleinfo,
    keyword = 'acute_myeloid_leukaemia',
    gene1 = 'TXNIP',
    gene2 = 'BRCA1',
    title = 'Acute myeloid leukaemia',
    titlesize = 24,
    axissize = 16)


```

Now, plot these together:

```{r plot_correlation, fig.width = 12, fig.height = 8}

  grid.arrange(p1, p2, ncol = 2, nrow = 1,
    top = textGrob('Correlation analysis', gp = gpar(fontsize = 24, fontface="bold")))

  grid.rect(gp = gpar(fill = NA))

```


# Session info

```{r}

sessionInfo()

```

